\documentclass[
  12pt      
  ,letterpaper
  ,center
  ,noupper
  ]{uconnthesis}
\ThesisLineSpacing{2}
\usepackage{graphicx}  % needed for figures
\usepackage{subeqnarray}
\usepackage{cases}
\usepackage{bm}        % for math
\usepackage{amssymb}   % for math
\usepackage{epstopdf}

\newcommand{\adag}{\ensuremath{a^{\dagger}}}
\newcommand{\bdag}{\ensuremath{b^{\dagger}}}
\newcommand{\cdag}{\ensuremath{c^{\dagger}}}
\newcommand{\ddagg}{\ensuremath{d^{\dagger}}}
\newcommand{\ket}[1]{\ensuremath{\left| #1 \right\rangle}}
\newcommand{\bra}[1]{\ensuremath{\left\langle #1 \right|}}
\newcommand{\braket}[2]{\ensuremath{\left\langle #1 \right|\left. #2\right\rangle}}
\newcommand{\mele}[3]{\ensuremath{\left\langle #1 \right|#2\left| #3\right\rangle}}
\newcommand{\dyad}[2]{{\ket{#1}\!\!\bra{#2}}}
\newcommand{\norm}[1]{\ensuremath{\left\| #1 \right\|}}
\newcommand{\abs}[1]{\ensuremath{\left| #1 \right|}}
\newcommand{\To}{\ensuremath{\rightarrow}}
\newcommand{\from}{\ensuremath{\leftarrow}}
\newcommand{\funE}{\ensuremath{\mathcal{E}}}
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\bea}{\begin{eqnarray}}
\newcommand{\eea}{\end{eqnarray}}
\newcommand{\mbf}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\eq}[1]{{(\ref{#1})}}
\newcommand{\commentout}[1]{{}}
\newcommand{\idx}{{l}}
\newcommand{\half}{{\hbox{$\frac{1}{2}$}}}
\newcommand{\Kappa}{{\cal K}}
\newcommand{\bE}{{\bf E}}
\newcommand{\br}{{\bf r}}
\newcommand{\cbE}{\boldsymbol{\mathbf{\cal E}}}
\newcommand{\cbP}{\boldsymbol{\mathbf{\cal P}}}
\newcommand{\dip}{{\cal D}}
\newcommand{\bdip}{\boldsymbol{\mathbf{\cal D}}}
\newcommand{\G}{{\sf G}}
\newcommand{\bd}{{\bf d}}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\ja}[1]{\red{#1}}
\newcommand{\js}[1]{\textcolor{blue}{\textst{#1}}}

\begin{document}
\bibliographystyle{prsty}

\abstract{
\begin{center}
{ ABSTRACT}
\end{center}
At high density, dipole-dipole interactions between the atoms may have a major impact on light propagation in a dense gas. We have developed a classical-electrodynamics simulation to study the cooperative response of a near-resonant gas to light. We take into account the motion of the radiators using classical trajectories, including collisions with the walls of the container and atom-atom collisions, and describe the transition from homogeneously broadened to inhomogeneously broadened phenomenology.
}

\title{Cooperative Effects in the Optical Response of Dense Atomic Gases}
\author{Yi Li}
\date{2016}
\authorspreviousdegreelong{
  B.S., University of Science and Technology of China, Hefei, China, 2007 \\
  M.S., University of Connecticut, 2010
  }
\authorspreviousdegreeshort{M.S.}

\MajorAdvisor{Juha Javanainen}
\AssociateAdvisorA{Robin C\^ot\'e}
\AssociateAdvisorB{William Stwalley}

\dedication{
Dedicated to my beloved parents and my wife Yixin
  }

\acknowledgements{
First of all, I would like to express my deepest gratitude to my advisor, Prof. Juha Javanainen. I feel amazingly fortunate to have an advisor who gave me so much insightful guidance on my research as well as on my future career. Without your priceless encouragement and support, the past eight years could not have been such an incredibly rewarding journey in my life that I will cherish forever.Ê\\
I am very grateful to my associate advisors Prof. William Stwalley and Prof. Robin C\^ot\'e. Your comments and criticisms are precious sources of inspirations. I greatly appreciate your support throughout this dissertation research.\\
I would like to thank Prof. Richard Jones and especially Dr. Igor Senderovich for your warm help in using the computing facilities to carry out the numerical simulations involved in this dissertation.\\
I would like to thank some other faculty members including Prof. Gerald Dunne, Prof. Gayanath Fernando, Prof. Philip Mannheim, Prof. Ronald Mallett, Prof. Alexander Kovner and Prof. Peter Schweitzer. I learned a lot on various branches of physics from your excellent courses.Ê\\
I am thankful to the staff members Kim Giard, Dawn Rawlinson and Micki Bellamy in the main office for your kind help. Thanks to Carol Guerra, Heather Osborne, Michael Rapposch and Prof. G.Brooks Jr for your assistance in my teaching experience. \\
I would like to thank my peer physics graduate students for the friendly community in the department. Thanks to Dr. Han Chen, Dr. Otim Odong, Dr. J. C. Sanders, Dr. Shaozhen Ma, Dr. Kuang He, Dr. Yuefeng Nie, Dr. Zhe Chen, Dr. Kun Fang, Dr. Hui Chen, Dr. Liang Dong, Kun Fu, Xugang He, Fu-Chang Sun, Dr. Zhihai Zhu, Zhiwei Zhang, Leland Aldridge, James Wells and others in the department.\\
I would like to thank a lot of friends that I met in these years at University of Connecticut. Thanks to Yige Zhao, Hui Huang, Xiguang Yan, Qian He, Luye Chang, Shuang Yin, Yun Zhang, Danjie Zhang for their friendship. In particular, I would like to mention a close friend of my family, Shan Hu, she graduated from University of Connecticut with a Ph.D. in 2012 and passed away unexpectedly in March 2016.  I and my family will always remember her.\\
Last but not least, I would like to thank my parents and wife Yixin Xu, for all the love from them throughout these years.
}

\maketitle
\frontmatter
\tableofcontents
\listoffigures
%\listoftables

\mainmatter

\input{chapter1_r1.tex}
\input{chapter2_r1.tex}
\input{chapter3_r1.tex}
\input{chapter4_r1.tex}
\input{chapter5_r1.tex}
\input{chapter6_r1.tex}




\bibliography{bib_summary}



% appendices: we put them after the references in what ever order
% we prefer
%\input{appendices/dsmethod.tex}
\end{document}

